#Reproducible Research: Course Project 2

Sima N
May 21, 2020 


##Synopsis
This report explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to identify when and where major storms and weather events occur in the United States, and what is the estimate of any fatalities, injuries, and property damage.

1 - which types of events are most harmful to population health?

2 - which types of events have the greatest economic consequences?

Results show that tornados are the most damaging weather type to population health; and floods are the most damaging weather event.

knitr::opts_chunk$set(warning=FALSE)

##Data Processing
##Loading the data into R

```{r, echo = TRUE}

#add libraries 
library(ggplot2)
library(dplyr)
library(knitr)


#download data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, destfile ='repdata-data-StormData.csv.bz2')

#read data
stormdata <- read.csv("repdata-data-StormData.csv.bz2")
head(stormdata)

```
#variables used in the analysis:

EVTYPE: Event Type (Tornados, Flood, ..)

FATALITIES: Number of Fatalities

INJURIES: Number of Injuries

PROGDMG: Property Damage

PROPDMGEXP: Units for Property Damage (magnitudes - K,B,M)

CROPDMG: Crop Damage

CROPDMGEXP: Units for Crop Damage (magnitudes - K,BM,B)
```{r, echo = TRUE}

varsNedeed <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm <- stormdata[varsNedeed]
dim(storm)

```
#Results

#Which events are most harmful to population Health?
#Fatalities Data

```{r, echo = TRUE}

fatalityData <- (subset(stormdata, FATALITIES > 0))[c(8, 23)]
fatalityGroupedData <- aggregate(FATALITIES ~ EVTYPE, data = fatalityData, FUN = "sum", na.rm = TRUE)
fatalityGroupedData <- fatalityGroupedData[order(fatalityGroupedData$FATALITIES, decreasing=TRUE), ]
fatalityGroupedData <- fatalityGroupedData[1:5, ]
fatalityGroupedData$EVTYPE <- factor(fatalityGroupedData$EVTYPE, levels=fatalityGroupedData$EVTYPE)
```

#Injuries Data
```{r, echo = TRUE}
injuryData <- (subset(stormdata, INJURIES > 0))[c(8, 24)]
injuryGroupedData <- aggregate(INJURIES ~ EVTYPE, data = injuryData, FUN = "sum", na.rm = TRUE)
injuryGroupedData <- injuryGroupedData[order(injuryGroupedData$INJURIES, decreasing=TRUE), ]
injuryGroupedData <- injuryGroupedData[1:5, ]
injuryGroupedData$EVTYPE <- factor(injuryGroupedData$EVTYPE, levels=injuryGroupedData$EVTYPE)
```

#ploting the results

```{r, echo = TRUE}

ggplot(fatalityGroupedData, aes(x=EVTYPE, y=FATALITIES)) +
  geom_bar(stat="identity") +
  xlab("Types") + 
  ylab("Fatalities") +
  ggtitle("Total Fatalities")



ggplot(injuryGroupedData, aes(x=EVTYPE, y=INJURIES)) +
  geom_bar(stat="identity") +
  xlab("Types") + 
  ylab("Injuries") +
  ggtitle("Total Injuries")
```


#Which type of Events have the greatest Economic consequences?

```{r, echo = TRUE}
 
#Convert economic impact to monitary value.
damageAmount <- function(amount, magnitude)
{
  returnAmount <- 0
  if (toupper(magnitude)[1]=="K")
  {
    returnAmount <- (amount * 1000)
  }
  if (toupper(magnitude)[1]=="M")
  {
    returnAmount <- (amount * 1000000)
  }
  if (toupper(magnitude)[1]=="B")
  {
    returnAmount <- (amount * 1000000000)
  }
  return(returnAmount)
}

damageData <- (subset(stormdata, PROPDMG > 0 | CROPDMG > 0))[c(8, 25, 26, 27, 28)]
damageData$DamageAmount <- ((mapply(damageAmount, damageData$PROPDMG, damageData$PROPDMGEXP)) +
                                  (mapply(damageAmount, damageData$CROPDMG, damageData$CROPDMGEXP)))
                          
damageGroupedData <- aggregate(DamageAmount ~ EVTYPE, data = damageData, FUN = "sum", na.rm = TRUE)
damageGroupedData <- damageGroupedData[order(damageGroupedData$DamageAmount, decreasing=TRUE), ]
damageGroupedData <- damageGroupedData[1:5, ]
damageGroupedData$EVTYPE <- factor(damageGroupedData$EVTYPE, levels=damageGroupedData$EVTYPE)

#Most harmful events 
head(damageGroupedData, 5)
```
                        